---
title: "4Host Normalisation & DEG analysis"
author: "R. Stam"
date: "3 February 2015"
output: html_document
---

Load libraries
```{r, message=FALSE, warning=FALSE, error=FALSE}
library("limma")
library("statmod")
```

Read in data
```{r, message=FALSE, warning=FALSE, error=FALSE}
setwd("/Users/rstam/Dropbox/4host_microarray/Array_Raw/sorted_probes")
files <- dir(pattern="*\\.txt$")
data <- read.maimages(files, source="agilent",columns=list(G = "gMedianSignal", Gb = "gBGMeanSignal"))
data.bg <-backgroundCorrect(data, method="normexp")

setwd("/Users/rstam/Dropbox/4host_microarray/Analyses/Remco")
```

Remove negative and low expressed
```{r}
neg95 <- apply(data.bg$E[data.bg$genes$ControlType==-1,],2,function(data.bg) quantile(data.bg ,p=0.95)) # 95th percentile of the negative control probes used as negative number
cutoff <- matrix(neg95,nrow(data.bg),ncol(data.bg),byrow=TRUE)
isexpr <- rowSums(data.bg$E > cutoff) >= 4 # Cutoff set at 4 arrays, meaning probe must be above BG in at least one timepoint
table(isexpr) # 18000 Pcap probes on the array with values higher than BG.
data.bg <- data.bg[data.bg$genes$ControlType==0 & isexpr,] # regular probes are 0 and should be in isexpr
```

Quantile normalisation
```{r}
data.q<-normalizeBetweenArrays(data.bg,method="quantile")
data.q.avg <- avereps(data.q, ID=data.q$genes$ProbeName)

# Plot graphs, normalisation looks fine
par(mfrow=c(1,2))
boxplot(data$E, las=2, col= c("pink", "pink","pink","pink","red","red","red","red",
                              "green","green","green", "green","magenta","magenta","magenta", "magenta"))
boxplot(data.q$E, las=2, col= c("pink", "pink","pink","pink","red","red","red","red",
                                "green","green","green", "green","magenta","magenta","magenta", "magenta"))

```


Build design matrix
```{r}
Ca <-c(1,1,1,1,0,0,0,0,0,0,0,0,0,0,0,0)
Cs <-c(0,0,0,0,1,1,1,1,0,0,0,0,0,0,0,0)
Nb <-c(0,0,0,0,0,0,0,0,1,1,1,1,0,0,0,0)
Sl <-c(0,0,0,0,0,0,0,0,0,0,0,0,1,1,1,1)
design <- cbind(files,Ca,Cs,Nb,Sl)
rownames(design)<-design[,1]
design<-as.matrix(design[,2:5])
class(design)<-"numeric"
```

Fit data and create individual contrasts to analyse
```{r}
fit <- lmFit(data.q.avg, design)
contrasts <- makeContrasts(Ca-Cs,Ca-Nb,Ca-Sl,Cs-Nb,Cs-Sl,Nb-Sl,levels=c("Ca","Cs","Nb","Sl"))
contrastsCa <- makeContrasts(Ca-Cs,Ca-Nb,Ca-Sl,levels=c("Ca","Cs","Nb","Sl"))
contrastsCs <- makeContrasts(Cs-Ca,Cs-Nb,Cs-Sl,levels=c("Ca","Cs","Nb","Sl"))
contrastsNb <- makeContrasts(Nb-Ca,Nb-Cs,Nb-Sl,levels=c("Ca","Cs","Nb","Sl"))
contrastsSl <- makeContrasts(Sl-Ca,Sl-Cs,Sl-Nb,levels=c("Ca","Cs","Nb","Sl"))
```

Calculate DEGs in the whole experiment
```{r}
options(digits=2)

fitAll <- contrasts.fit(fit, contrasts)
fitAll <- eBayes(fitAll)
results <- decideTests(fitAll)
summary(results)
```

Write out table with p-values for all DEGs and table for all genes.
These files can be used to extract gene names for DEGs and also to see how significantly changing a gene is.
```{r}
topAll <- topTable(fitAll, adjust="BH", number = 3000, p.value = 0.05)
topAll_table <- topAll[,c(5,12:15)]
rownames(topAll_table) <- NULL
#write.table(topAll_table, file="2-1_4host_DEGs.txt", quote = FALSE, sep="\t" )

All <- topTable(fitAll, adjust="BH", number = 200000)
All_table <- All[,c(5,12:15)]
rownames(All_table) <- NULL
#write.table(All_table, file="2-2_4host_all.txt", quote = FALSE, sep="\t" )
```


Draw Venn diagrams to show the differently expressed genes between the species
```{r}
par(mfrow=c(2,2))
fitCa <- contrasts.fit(fit, contrastsCa)
fitCa <- eBayes(fitCa)
resultsCa <- decideTests(fitCa)
summary(resultsCa)
vennDiagram(resultsCa)

fitCs <- contrasts.fit(fit, contrastsCs)
fitCs <- eBayes(fitCs)
resultsCs <- decideTests(fitCs)
summary(resultsCs)
vennDiagram(resultsCs)

fitNb <- contrasts.fit(fit, contrastsNb)
fitNb <- eBayes(fitNb)
resultsNb <- decideTests(fitNb)
summary(resultsNb)
vennDiagram(resultsNb)

fitSl <- contrasts.fit(fit, contrastsSl)
fitSl <- eBayes(fitSl)
resultsSl <- decideTests(fitSl)
summary(resultsSl)
vennDiagram(resultsSl)
```

Turn data into dataframe for follow up.
```{r}
#Make df
data.q.frame <- as.data.frame.EList(data.q.avg, row.names = NULL, )
rownames(data.q.frame) <- data.q.frame$SystematicName
data.q.frame <- data.q.frame[,6:21]

# Calculate means
data.q.frame$mCa <- rowMeans(subset(data.q.frame, select = c("Ca_1_sorted", "Ca_2_sorted", "Ca_3_sorted", "Ca_4_sorted")), na.rm = TRUE)
data.q.frame$mCs <- rowMeans(subset(data.q.frame, select = c("Cs_1_sorted", "Cs_2_sorted", "Cs_3_sorted", "Cs_4_sorted")), na.rm = TRUE)
data.q.frame$mNb <- rowMeans(subset(data.q.frame, select = c("Nb_1_sorted", "Nb_2_sorted", "Nb_3_sorted", "Nb_4_sorted")), na.rm = TRUE)
data.q.frame$mSl <- rowMeans(subset(data.q.frame, select = c("Sl_1_sorted", "Sl_2_sorted", "Sl_3_sorted", "Sl_4_sorted")), na.rm = TRUE)

#Subset tables (all values, means)
data.means <- data.q.frame[,17:20]
data <- data.q.frame[,1:16]

#write.table(data, file="2_Normalised_data_all.txt", quote = FALSE, sep="\t" )
#write.table(data.means, file="2_Normalised_data_means.txt", quote = FALSE, sep="\t" )

```

Plot Barplots for four markergenes. If normalisation worked, they should all show similar levels in each host, but different levels when compared to each other: Tub and Hmp1 high, Cdc14 and Npp1 low.
```{r}
##MULTIPLE BARPLOTS: CONTROLS
sterr <- function(x) sd(x, na.rm = TRUE)/sqrt(length(x))
x <- list(1:4, 5:8, 8:12, 13:16)

## Get data for Hmp1 and tubulin and other 'controls'
par(mfrow=c(2,2))

Hmp1 <- do.call(cbind, lapply(x, function(i) rowMeans(data["Phyca11_8692",i])))
Hmp1Sd <- do.call(cbind, lapply(x, function(i) sd(data[1,i])))
Hmp1Ser <- do.call(cbind, lapply(x, function(i) sterr(data[1,i])))

bary <- barplot(Hmp1, ylab = "N exp", ylim = c(0,20), main = "Hmp1", col = "lightgrey")
labels <- paste(c("pepper","cucumber","benthamiana","tomato"))
text(c(0.75,2,3.25,4.5), par("usr")[3] - 1, srt = 45, adj = 1,
     labels = labels, xpd = TRUE)
arrows(bary ,Hmp1+Hmp1Ser, bary, Hmp1, angle=90, code=1, length = 0.05)
arrows(bary ,Hmp1-Hmp1Ser, bary, Hmp1, angle=90, code=1, length = 0.05)


Tub <- do.call(cbind, lapply(x, function(i) rowMeans(data["Phyca11_576734",i])))
TubSd <- do.call(cbind, lapply(x, function(i) sd(data[1,i])))
TubSer <- do.call(cbind, lapply(x, function(i) sterr(data[1,i])))

bary <- barplot(Tub, ylab = "N exp", ylim = c(0,15), main = "Tubulin", col = "lightgrey")
labels <- paste(c("pepper","cucumber","benthamiana","tomato"))
text(c(0.75,2,3.25,4.5), par("usr")[3] - 1, srt = 45, adj = 1,
     labels = labels, xpd = TRUE)
arrows(bary ,Tub+TubSer, bary, Tub, angle=90, code=1, length = 0.05)
arrows(bary ,Tub-TubSer, bary, Tub, angle=90, code=1, length = 0.05)

Npp1 <- do.call(cbind, lapply(x, function(i) rowMeans(data["Phyca11_11951",i])))
Npp1Sd <- do.call(cbind, lapply(x, function(i) sd(data[1,i])))
Npp1Ser <- do.call(cbind, lapply(x, function(i) sterr(data[1,i])))

bary <- barplot(Npp1, ylab = "N exp", ylim = c(0,10), main = "Npp1", col = "lightgrey")
labels <- paste(c("pepper","cucumber","benthamiana","tomato"))
text(c(0.75,2,3.25,4.5), par("usr")[3] - 1, srt = 45, adj = 1,
     labels = labels, xpd = TRUE)
arrows(bary ,Npp1+Npp1Ser, bary, Npp1, angle=90, code=1, length = 0.05)
arrows(bary ,Npp1-Npp1Ser, bary, Npp1, angle=90, code=1, length = 0.05)

Cdc14 <- do.call(cbind, lapply(x, function(i) rowMeans(data["Phyca11_510939",i])))
Cdc14Sd <- do.call(cbind, lapply(x, function(i) sd(data[1,i])))
Cdc14Ser <- do.call(cbind, lapply(x, function(i) sterr(data[1,i])))

bary <- barplot(Cdc14, ylab = "N exp", ylim = c(0,10), main = "Cdc14", col = "lightgrey")
labels <- paste(c("pepper","cucumber","benthamiana","tomato"))
text(c(0.75,2,3.25,4.5), par("usr")[3] - 1, srt = 45, adj = 1,
     labels = labels, xpd = TRUE)
arrows(bary ,Cdc14+Cdc14Ser, bary, Cdc14, angle=90, code=1, length = 0.05)
arrows(bary ,Cdc14-Cdc14Ser, bary, Cdc14, angle=90, code=1, length = 0.05)
```


```{r}
## WRITE OUTPUT TABLE
data.means.DEG <- data.means[topAll_table[,1],]
write.table(data.means.DEG, file="2_Normalised_data_DEGs.txt", quote=FALSE, sep="\t")
```